---
title: "NEON site vegetation structure"
author: "Lizbeth G Amador"
date: "2023-06-19"
output: pdf_document
---

```{r, libraries, echo=FALSE}
require(neonUtilities, quietly = TRUE)
require(neonOS, quietly = TRUE)
require(tidyverse, quietly = TRUE)
```


Presets
```{r, preset}
#keep strings as character 
options(stringsAsFactors = F)
```


# Data Acquisition  
Read in 'Vegetation Structure' data product form the NEON data portal 
```{r}
vegs = loadByProduct(
  dpID = "DP1.10098.001", #Data product ID - Required
  site = c("RMNP", "OSBS", "ABBY", "GUAN", "MLBS"), 
  startdate = "2018-01", enddate = "2018-12", #Dates - Required
  check.size = F, #will notify size before downloading - answer prompt to proceed!
  release = 'RELEASE-2023' #most recent version
)
```

Looking at the variables and table descriptions for each table; multiple 
dataframes exist in the `vegs` object (includes metadata tables).
```{r, variables and metadata}
#peeking at the variable table -- same function with pre-downloaded data
View(vegs$variables_10098)
#write.csv(vegs$variables_10098, file="NEON_VegStructure_VariableList.csv", row.names = FALSE)
#Gives description for each table
View(vegs$readme_10098)
```
 
 
Saving dataframes from object tables  
```{r, slecting tables}
#woody plant biomass & productivity 
woody = vegs$vst_apparentindividual
#write.csv(woody, "NEON_Woody_VegStructure_selectsites2018.csv", row.names = FALSE)

#non-woody plant (non-herbaceous perennial) biomass & productivity 
nonwoody = vegs$`vst_non-woody`
#write.csv(nonwoody, "NEON_NonWoody_VegStructure_selectsites2018.csv", row.names = FALSE)

#table for taxonomic info
maptag = vegs$vst_mappingandtagging
```


Looking into the `woody` data
```{r}
#structure of variables
str(woody)
#dataframe dimension
dim(woody)
#name of columns
names(woody)
```

Looking into `nonwoody`  
```{r}
#structure of variables
str(nonwoody)
#dataframe dimensions
dim(nonwoody)
#name of columns
names(nonwoody)
```

Looking into `maptag`  
```{r}
#structure of variables
str(maptag)
#dataframe dimensions
dim(maptag)
#name of columns
names(maptag)
```

# Cleaning
Remove duplicates in the data & see what common fields `woody` and `maptag` share
```{r, remove duplicates}
#drop IUD as that will be unique for duplicate records
woody_noUID = select(woody, -(uid))
maptag_noUID = select(maptag, -(uid))

#remove duplicates 
woody_noD = distinct(woody_noUID)
dim(woody)
dim(woody_noD)

maptag_noD = distinct(maptag_noUID)
dim(maptag)
dim(maptag_noD)

#pull overlapping column names
intersect(names(woody_noD), names(maptag_noD))
```


Some column values have unique values and we want to keep those distinct and not 
join them so we will rename them.
```{r, renaming columns}
maptag_noD = rename(maptag_noD, 
                    date_map=date, remarks_map=remarks, 
                    recordedBy_map=recordedBy, measuredBy_map=measuredBy, 
                    dataQF_map=dataQF, publicationDate_map=publicationDate)
```


# Join Dataframes  
Now we can join the two data frames on all the variables with the same name. We 
use a `left_join()` from the `dpylr` package in `tidyverse` because we want to 
match all the rows from the "left" (first) dataframe to any rows that also occur 
in the "right" (second) dataframe.  
```{r}
wood_vegstr = left_join(maptag_noD, woody_noD,
                        by=c("namedLocation", "eventID", "domainID", "siteID", 
                             "plotID", "individualID", "release"))
```

```{r, data specs}
dim(wood_vegstr)
names(wood_vegstr)
```


Let's save the cleaned joined dataset
```{r, export}
#exports as csv file
write.csv(wood_vegstr, file="Output_202306/Wood_VegStructure_selectSites_2018_noDups.csv", row.names = FALSE)
#saves the object as an .RDdata file (compressed dataframe)
save(wood_vegstr, file = "Output_202306/wood_vegstr2018.RData")
```


Now let's just filter for a specific site (for in-person week) and see the 
amount of plots in there.

What do the plots look like?
```{r}
sort(unique(wood_vegstr[wood_vegstr$siteID=="MLBS",]$plotID))
#unique(wood_vegstr[wood_vegstr$siteID=="MLBS",]$subplotID)
#unique(wood_vegstr[wood_vegstr$siteID=="MLBS",]$nestedSubplotID)
names(wood_vegstr)
```

Subset for MLBS only
```{r}
mlbs_wood = wood_vegstr %>%
  filter(siteID=="MLBS")
unique(mlbs_wood$siteID)
```

# Structural diversity metrics: basal area  

We will want to calculate basal area using the `stemDiameter` as a metric of 
structural diversity as a relative abundance for each species in a plot/site
```{r}
unique(mlbs_wood$stemDiameter)
mlbs_wood %>%
  count(stemDiameter)

mlbs_wood %>%
  count(taxonID)
```

Frequency of stem diameter
```{r}
hist(mlbs_wood$stemDiameter, xlab="Stem Diamter (cm)",
     main="MLBS Stem Diameter")
```


Checking stats for each plot with respect to stemDiamter  
```{r}
library(mosaic, quietly = TRUE)

favstats(stemDiameter~plotID, data=mlbs_wood)
```


Filtering for stemDiamter for greater than 10cm  
```{r}
mlbs_wood_lim = mlbs_wood %>%
  filter(stemDiameter>=10)

favstats(stemDiameter~plotID, data=mlbs_wood_lim)
```


Making a boxplot
```{r}
png(filename = "Output_202306/MLBS_StemDiameter.png", width=1050, height=800, units=)
boxplot(mlbs_wood_lim$stemDiameter~mlbs_wood_lim$plotID, 
        main="Plot Comparison at MLBS",
        xlab="Plot ID",
        ylab = "Stem Diamter (cm)")
dev.off()
```
















